---
title: "Normalization and Clustering of scRNA-seq data of Naka et al."
author: "Davide Risso"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
editor_options:
  chunk_output_type: inline
---

```{r options, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=FALSE)
# NMF::nmf.options(grid.patch=TRUE) #get rid of blank page in heatmap pdf

library(SummarizedExperiment)
library(TENxGenomics)
library(scater)
library(scone)
library(clusterExperiment)
library(ggplot2)
library(magrittr)
library(cowplot)
library(Rtsne)
library(zinbwave)
library(uwot)
library(Seurat)
library(clustree)
library(ggridges)
library(scmap)
```

# Introduction

In this document we show the code used for the analysis of the single-cell RNA-seq data generated in Naka et al.

The goal is to characterize the transcriptional diversity of Sst cells.

We used the 10x genomics preprocessing pipeline, and we'll start from a matrix of
UMI counts that need to be normalized and clustered. 

The data folder includes both raw counts and processed data for reproducibility.

To read in the data, we use the [TENxGenomics package](https://github.com/mtmorgan/TENxGenomics) by Martin Morgan. This is not necessary, a simple function that reads in a HDF5 file would be sufficient, but this package provides a nice interface to the `SummarizedExperiment` class.

```{r readin}
tdt <- matrixSummarizedExperiment("data/tdt/filtered_gene_bc_matrices_h5.h5")
gfptdt <- matrixSummarizedExperiment("data/gfptdt/filtered_gene_bc_matrices_h5.h5")
tdt1 <- matrixSummarizedExperiment("data/tdt1/filtered_gene_bc_matrices_h5.h5")
tdt2 <- matrixSummarizedExperiment("data/tdt2/filtered_gene_bc_matrices_h5.h5")

se <- cbind(tdt1, tdt2, tdt, gfptdt)
se
colData(se)$Batch <- as.factor(c(rep("TdT+1", NCOL(tdt1)),
                                 rep("TdT+2", NCOL(tdt2)),
                                 rep("TdT+3", NCOL(tdt)),
                                 rep("TdT+GFP+", NCOL(gfptdt))))
colData(se)
```

We processed four batches for a total of `r NCOL(se)` cells.

# Exploratory Data Analysis

Looking at library sizes and detected genes.

```{r eda}
hist(colSums(assay(se)), breaks=30, 
     xlab='Number of UMI', main="Number of UMI per sample")
hist(colSums(assay(se)>0)/nrow(se), breaks=30, 
     xlab="Proportion of detected genes", main="Proportion of detected genes")

hist(colSums(assay(se)>0), breaks=30, 
     xlab="Number of detected genes", main="Number of detected genes")

hist(colMeans(assay(se)==0), breaks=30, 
     xlab="Proportion of zeros", main="Proportion of zeros")

boxplot(colSums(assay(se))~colData(se)$Batch, main="Number of UMI per sample")
boxplot(colSums(assay(se)>0)/nrow(se)~colData(se)$Batch, main="Proportion of detected genes")
boxplot(colMeans(assay(se)==0)~colData(se)$Batch, main="Proportion of zeros")

simple <- assay(se)[rowSums(assay(se))>10,]
simple <- SUM_FN(simple)

library(rARPACK)
fastpca <- function(expr, scale=FALSE, k=50) {
  svd_raw <- svds(scale(t(expr), center=TRUE, scale=scale), k=k, nu=k, nv=0)
  pc_raw <- svd_raw$u %*% diag(svd_raw$d[1:k])
  return(pc_raw)
}
pca <- fastpca(log1p(simple), scale = TRUE)
```

We use `scater` to explore a set of QC metrics.

```{r scater}
sce <- as(se, "SingleCellExperiment")
assayNames(sce) <- "counts"
sce <- calculateQCMetrics(sce)

ribo_idx <- grep("^Rpl", rowData(se)[,2])
mito_idx <- grep("^Mt", rowData(se)[,2])
ribo_pct <- colSums(assay(se)[ribo_idx,])/colSums(assay(se)) * 100
mito_pct <- colSums(assay(se)[mito_idx,])/colSums(assay(se)) * 100
tdtomato <- assay(se)["tdTomato",]
gfp_exp <- assay(se)["GFP",]

Xist <- assay(se)["ENSMUSG00000086503",]

qc <- as.matrix(data.frame(colData(sce)[,c(7, 9:13)], mito_pct = mito_pct, ribo_pct = ribo_pct,
                           Xist = Xist, tdtomato = tdtomato, gfp = gfp_exp))
qcpca <- prcomp(qc, scale. = TRUE)
plot(qcpca$x, col=bigPalette[1], pch=19, main="QC PCA")

cors <- lapply(1:3, function(i) abs(cor(pca[,i], qc, method="spearman")))
cors <- unlist(cors)
bars <- data.frame(AbsoluteCorrelation=cors,
                   QC=factor(rep(colnames(qc), 3), levels=colnames(qc)),
                   Dimension=as.factor(rep(paste0("PC", 1:3), each=ncol(qc))))

bars %>%
  ggplot(aes(Dimension, AbsoluteCorrelation, group=QC, fill=QC)) +
  geom_bar(stat="identity", position='dodge') +
  scale_fill_manual(values=bigPalette) + ylim(0, 1) + 
  ggtitle("Correlation between QC and expression PCA")

fig_data <- data.frame(PC1 = pca[,1], PC2 = pca[,2], qc, colData(se),
                   QPC1=qcpca$x[,1], QPC2=qcpca$x[,2])

fig_pca <- ggplot(fig_data, aes(x = PC1, y = PC2, color = log10_total_counts)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")
fig_pca

fig_qpca <- ggplot(fig_data, aes(x = QPC1, y = QPC2, color = log10_total_counts)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")
fig_qpca

fig_pca <- ggplot(fig_data, aes(x = PC1, y = PC2, color = ribo_pct)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")
fig_pca

fig_qpca <- ggplot(fig_data, aes(x = QPC1, y = QPC2, color = ribo_pct)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")
fig_qpca

fig_pca <- ggplot(fig_data, aes(x = PC1, y = PC2, color = tdtomato)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")
fig_pca

fig_qpca <- ggplot(fig_data, aes(x = QPC1, y = QPC2, color = tdtomato)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")
fig_qpca

fig_pca <- ggplot(fig_data, aes(x = PC1, y = PC2, color = Batch)) +
  geom_point() + scale_color_brewer(palette = "Set1")
fig_pca

heatmap(cor(qc, method = "spearman"))
```

# Filtering

We use `scone` to filter out lowly-expressed genes and low-quality cells.

```{r filtering}
colData(se) <- colData(sce)

# select expressed genes only
se <- se[rowSums(assay(se))>0,]

# select common genes
num_reads <- quantile(assay(se)[assay(se) > 0])[4]
num_cells = 0.25*ncol(se)
is_common = rowSums(assay(se) >= num_reads ) >= num_cells
table(is_common)

# select housekeeping genes
data("housekeeping")
hk <- stringr::str_to_title(housekeeping[,1])
hk_idx <- which(rowData(se)$Symbol %in% hk)

mfilt <- metric_sample_filter(assay(se),
                             nreads = colData(se)$total_counts,
                             gene_filter = is_common,
                             pos_controls = hk_idx,
                             hard_nreads = 2000,
                             zcut = 3, mixture = FALSE,
                             plot = TRUE)

plot(pca, pch=19, col=bigPalette[as.numeric(mfilt$filtered_breadth)+1],
     main = "PCA Filtered on transcriptome 'breadth'")
plot(pca, pch=19, col=bigPalette[as.numeric(mfilt$filtered_fnr)+1],
     main = "PCA Filtered on FNR AUC")

plot(qcpca$x, pch=19, col=bigPalette[as.numeric(mfilt$filtered_breadth)+1],
     main = "QPCA Filtered on transcriptome 'breadth'")
plot(qcpca$x, pch=19, col=bigPalette[as.numeric(mfilt$filtered_fnr)+1],
     main = "QPCA Filtered on FNR AUC")

table(mfilt$filtered_nreads, mfilt$filtered_fnr)
filter_cell <- !apply(simplify2array(mfilt[!is.na(mfilt)]),1,any)

plot(qcpca$x, pch=19, col=bigPalette[as.numeric(filter_cell)+1],
     main = "PCA Filtered on FNR AUC")

table(filter_cell, se$Batch)

# Final Gene Filtering: Highly expressed in at least 5 cells
num_reads <- quantile(assay(se)[assay(se) > 0])[4]
num_cells = 5
is_quality = rowSums(assay(se) >= num_reads ) >= num_cells

filtered <- se[is_quality, filter_cell]

qc <- qc[colnames(filtered),]
```

After filtering, we retain `r NROW(filtered)` genes and `r NCOL(filtered)` cells.

```{r save_filtered}
save(filtered, file="data/combined_filtered_all_batches.rda")
save(qc, file="data/combined_filtered_qc.rda")
```

# Check for batch effects

```{r batch}
# check for batch effects
pca <- fastpca(log1p(SUM_FN(assay(filtered))), scale = TRUE, k=50)
tsne_data <- Rtsne(pca, pca=FALSE, max_iter = 5000)
qcpca <- prcomp(qc, scale. = TRUE)

plot(pca, pch=19, col=bigPalette[colData(filtered)$Batch],
     main = "PCA Color-coded by batch")
plot(qcpca$x, pch=19, col=bigPalette[colData(filtered)$Batch],
     main = "QPCA Color-coded by batch")

boxplot(pca[,1] ~ colData(filtered)$Batch)
boxplot(pca[,2] ~ colData(filtered)$Batch)

plot(tsne_data$Y, pch=19, col=bigPalette[colData(filtered)$Batch])
legend("bottomleft", levels(filtered$Batch), fill=bigPalette)

idx <- colnames(filtered)
fig_data <- data.frame(PC1=pca[,1], PC2=pca[,2], colData(filtered),
                       TSNE1=tsne_data$Y[,1], TSNE2=tsne_data$Y[,2])

ggplot(fig_data, aes(x = PC1, y = PC2, color = log10_total_counts)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")
```

# Normalization

## SCONE

We use the `scone` package to perform and compare several normalization methods. 

The normalized data will be use only for exploratory purposes, as the dimensionality reduction (see next step) carries its own normalization.

Note: since this step is computationally heavy, we have saved the results in the `data/` folder and by default running this document will load the pre-computed results. To re-run the scone function, change the following line to `run_scone <- TRUE`.

```{r scone}
run_scone <- FALSE

## select positive controls (from Allen paper)
gabaergic <- c("Crispld2", "Cxcl14", "Tpm2", "Itih5", "Cox6a2",
               "Cxcl14", "Car4", "Tac2", "Reln", "Npy2r", "Tnfaip8l3", "Cadps2", "2310042E22Rik", "Egln3", "Tpd52l1", "Megf10", "Aebp1", "Slc18a3", "Pvrl4", "Nrp1", "Sema5b", "Pcdh15", "Phlda1", "Bcar3", "Mab21l1", "Pbx3", "Nrp1", "Crh", "Lamp5", "Tnfaip8l3", "Atp6ap1l", "Gabrd", "Npy", "Pde11a", "Has2", "Krt12", "2310042E22Rik", "Ndst4", "Tnnt1", "Reln", "Mpped1", "Pde1a", "Pcdh18", "4921511H03Rik", "Rgs12", "Cd34", "Egln3", "Thsd7b", "Reln", "Krt12", "2310042E22Rik", "Ndst4", "Lhx6", "Cdca7", "Myo5b", "Pdlim3", "Efcab6", "Tnnt1", "Cryab", "Nfib", "Kit", "Lamp5", "Cd34", "Sln", "Npy", "Col14a1", "Rasl11a", "Frmd7", "Edn3", "Tnfaip8l3", "Frem1", "Scml4", "St6galnac5", "Pbx3", "Fam107a", "Krt73", "Cyb5r2", "Kcns3", "Rasl11a", "Timp3", "Adamts18", "Ano3", "Rasl11a", "Tnni3k", "Bdnf", "Gabrg1", "Spp1", "Nr4a2", "Hspb3", "Nts", "Myh8", "Chrna2", "Glra3", "Kit", "Ppapdc1a", "Nr2f2", "Gfra2", "Myh1", "Myh4", "Myh13", "Grm3", "Il1rapl2", "Tnni3k", "Cartpt", "Chrna2", "Tmem90a", "Nr2f2", "Myh4", "Myh13", "Myh1", "C1qtnf7", "4930503E14Rik", "Gm5622", "Gm8267", "Efemp1", "Grm3", "Cartpt", "Crh", "Pla2g4a", "Sgpp2", "Trpv6", "Klhl14", "Chrnb3", "Grm3", "Ano3", "Fam5c", "Htr2a", "Irs4", "Tacr1", "Nos1", "Bace2", "Ccdc109b", "Dnase1l3", "Gpr126", "Gpr151", "Gstm6", "Hcrtr1", "Htr7", "Krt18", "Insl6", "Ndst4", "Sit1", "9430021M05Rik", "Gabrg1", "Cdca7", "Ndst4", "Gpr88", "Man1a", "Th", "Calca", "Nell1", "Fxyd6", "Tac1", "Tacr3", "Ednrb", "Thsd7a", "Fosb", "Il1rapl2", "Tll1", "Col25a1", "Calb1", "Kit", "Etv1", "Bdnf", "Aloxe3", "Nt5e", "Cntnap5b", "Thsd7a", "Snca", "Cacna2d3", "Kit", "Pcdh8", "Olfm3", "Pdlim3", "Nfib", "Krt12", "Cdca7", "Thsd7a", "Gm1051", "Caln1", "Calb1", "Il1rapl2", "Col25a1", "Tac1", "Tpbg", "Gpx3", "Akr1c18", "Ntf3", "Lrrc61", "Tac1", "Tusc5", "Fign", "Il1rapl2", "Acsbg1", "Tac1", "Il1rapl2")

glutamatergic <- c("Rspo1", "Inhba", "Sparcl1", "Pde1a", "Lmo3", "Rorb", "Whrn", "Rspo1", "Endou", "Tmem215", "Rorb", "Whrn", "Pde1a", "Endou", "Plb1", "Aldh1l1", "Etv1", "Deptor", "Rorb", "Whrn", "Cpne7", "Scnn1a", "Endou", "Rspo1", "Whrn", "Stard8", "Inhba", "Wfs1", "Otof", "Enpp2", "Palmd", "Rgs8", "Dgkb", "Adamts2", "Fst", "Matn2", "Cdh13", "Dgkb", "Otof", "Pacsin2", "Myl4", "Deptor", "Tmem91", "Whrn", "Il1rapl2", "Hsd11b1", "Nnat", "Syt17", "Myl4", "D430036J16Rik", "Deptor", "Arhgap25", "Cpne2", "Il1rapl2", "Foxo1", "Deptor", "Myl4", "Aldoc", "Arhgap25", "Il1rapl2", "Hhatl", "Itga7", "Myl4", "Arhgap25", "Man1a", "Penk", "Anxa11", "Cd7", "Cited1", "Lipg", "Nnat", "Ptchd2", "Sorcs3", "Inhba", "C1ql3", "Pter", "Acvr1c", "Fst", "Prss22", "Traip", "Ptprk", "Il1rapl2", "Ctgf", "Tnmd", "Nxph4", "Col24a1", "Trh", "Fam46a", "Ngf", "Moxd1", "Ndrg1", "Tmem40", "Cplx3", "Cidea", "Inpp4b", "Lman1l", "Igsf3", "Ctgf", "Nxph4", "Cidea", "Ly6g6e", "Sla", "Cplx3", "Clic5", "Lman1l", "Gpr126", "Arhgap25", "Foxp2", "Ly6d", "Rprm", "Ifitm2", "Ctxn3", "Crym", "Foxp2", "Gabra5", "Rprm", "Plekhb1", "Slc6a11", "Crym", "Chrnb3", "Scml2", "Ngb", "Klk8", "Plac9", "Ddit4l", "Fam84b", "Qrfpr", "Samd3", "Stac", "Ddit4l", "2310042E22Rik", "Kcns3", "Mc4r", "Coro6", "Sema3c", "Kctd8", "Crym", "Fam84b", "Ptgfr", "Depdc7", "Qrfpr", "Col6a1", "Syt17", "2310042E22Rik", "Crym", "Man1a", "Fam84b", "Ctxn3")

nonneruonal <- c("F3", "Rorb", "Acsbg1", "Slc39a12", "Ntsr2", "Plcd4", "Gja1", "Gjb6", "Cbs", "Chrdl1", "Prodh", "Mlc1", "Acsl6", "Slc4a4", "Gabrg1", "Cxcl14", "Slco1c1", "Vcam1", "Ednrb", "Scrg1", "Bcan", "Cspg4", "Pcdh15", "Gria3", "Cacng4", "E130309F12Rik", "Vcan", "Ednrb", "Scrg1", "Bcan", "Gpr17", "Brca1", "Rnf122", "Mbp", "Zcchc12", "Enpp6", "Kif19a", "Enpp6", "Dct", "Tmeff2", "Gpr17", "1700040N02Rik", "1810041L15Rik", "St18", "Vcan", "Bcan", "9530059O14Rik", "Cldn11", "1700047M11Rik", "Mbp", "Mog", "Aspa", "Mobp", "Gpr37", "Ppp1r14a", "Gjb1", "Tmeff2", "St18", "Cldn11", "1700047M11Rik", "Kctd13", "Cntn2", "Eml1", " A530088E08Rik", "Cx3cr1", "C1qb", "Cd53", "Csf1r", "Itgam", "Abi3", "C1qa", "Aif1", "Trem2", "P2ry13", "Tmem119", "C1qc", "Cd14", "Fcgr3", "Gpr34", "Inpp5d", "Nckap1l", "Mpeg1", "Siglech", "Susd3", "Hk2", "Ly86", "Sparc", "Fli1", "Tbc1d4", "AI467606", "Exosc7", "Eltd1", "Fas", "Hmgcs2", "Nostrin", "Paqr5", "Slc16a4", "Id1", "Ptprb", "Cd93", "Sparc", "Fli1", "Ly6a", "Ly6c1", "Ly6c2", "Flt1", "Pglyrp1", "Slco1a4", "Ifitm3", "Abcb1a", "Ahnak", "Bgn", "Nupr1", "Casq2", "Mylk", "Gprc5c", "Slc38a11", "Slc6a20a", "Pcolce", "Vtn", "Cnn2", "Nid1", "Gpr30", "Higd1b", "Ifitm1", "P2ry14", "Serping1", "Sparc", "Fli1", "Cald1", " Abcb1a", "Flt1", "Ly6a", "Ly6c1", "Ly6c2", "Pglyrp1", "Slco1a4", "Ahnak")

poscon <- intersect(c(gabaergic, glutamatergic, nonneruonal), rowData(filtered)$Symbol)
rowData(filtered)$poscon <- (rowData(filtered)$Symbol %in% poscon)

## select negative controls (housekeeping)
hk <- intersect(hk, rowData(filtered)$Symbol)
set.seed(42)
negconeval <- sample(hk, length(poscon))
negconruv <- setdiff(hk, negconeval)
rowData(filtered)$negcon_eval <- (rowData(filtered)$Symbol %in% negconeval)
rowData(filtered)$negcon_ruv <- (rowData(filtered)$Symbol %in% negconruv)

names <- colnames(qc)[1:8]
if(run_scone) {
  library(BiocParallel)
  library(doParallel)
  registerDoParallel(7)
  register(DoparParam())
  
  scone_obj <- SconeExperiment(filtered,
                               which_qc = which(colnames(colData(filtered)) %in% names),
                               which_negconruv = which(colnames(rowData(filtered)) == "negcon_ruv"),
                               which_negconeval = which(colnames(rowData(filtered)) == "negcon_eval"),
                               which_poscon = which(colnames(rowData(filtered)) == "poscon"),
                               which_batch = which(colnames(colData(filtered)) == "Batch"))
  
  scone_obj <- scone(scone_obj,
                     scaling = list(none=identity, sum = SUM_FN, tmm = TMM_FN,
                                    fq = FQT_FN, deseq = DESEQ_FN),
                     zero = "postadjust",
                     adjust_batch = "no")
  save(scone_obj, file="data/combined_scone_res_all_batches.rda")
} else {
  load("data/combined_scone_res_all_batches.rda")
}

norm <- get_normalized(scone_obj, "none,fq,ruv_k=5,no_bio,no_batch")
lognorm <- log1p(norm)
```

## ZINB-WaVE

We use the `zinbwave` Bioconductor package for the dimensionality reduction of the data.

In particular, we use `K=10` dimensions to describe the data and `ribo_pct`, `total features`, and `batch` as covariates.

Note: since this step is computationally heavy, we have saved the results in the `data/` folder and by default running this document will load the pre-computed results. To re-run the zinbwave function, change the following line to `run_zinbwave <- TRUE`.


```{r zinbwave}
run_zinbwave <- FALSE

vars <- rowVars(lognorm)
names(vars) <- rownames(filtered)
vars <- sort(vars, decreasing = TRUE)

colData(filtered)$ribo_pct <- qc[,"ribo_pct"]

if(run_zinbwave) {
library(BiocParallel)
library(doParallel)
registerDoParallel(7)
register(DoparParam())

zinb <- zinbFit(filtered[names(vars)[1:1000],], 
                X = "~ Batch + ribo_pct + log10_total_features", 
                K=10, epsilon=1000)
save(zinb, file="data/combined_zinbwave_all_batches.rda")
} else {
  load("data/combined_zinbwave_all_batches.rda")
}

W <- getW(zinb)
rownames(W) <- paste0("Cell", seq_len(NROW(W)))
colnames(W) <- paste0("W", 1:10)

tsne_zinb <- Rtsne(W, pca = FALSE, max_iter=5000)
umap_zinb <- umap(W, n_neighbors = 15)
```

```{r zinbwave_plot, dependson="zinbwave"}
plot(W[,1:2], pch=19, col=bigPalette[colData(filtered)$Batch])

plot(tsne_zinb$Y, pch=19, col=bigPalette[colData(filtered)$Batch])
legend("topright", levels(colData(filtered)$Batch), fill=bigPalette)

idx <- colnames(filtered)
fig_data <- data.frame(W1 = W[,1], W2 = W[,2], colData(filtered),
                       TSNE1=tsne_zinb$Y[,1], TSNE2=tsne_zinb$Y[,2],
                       UMAP1=umap_zinb[,1],
                       UMAP2=umap_zinb[,2]
                      )

ggplot(fig_data, aes(x = W1, y = W2, color = log10_total_counts)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")

ggplot(fig_data, aes(x = W1, y = W2, color = log10_total_features_by_counts)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")

ggplot(fig_data, aes(x = W1, y = W2, color = ribo_pct)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")

ggplot(fig_data, aes(x = TSNE1, y = TSNE2, color = log10_total_counts)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")

ggplot(fig_data, aes(x = TSNE1, y = TSNE2, color = log10_total_features_by_counts)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")

ggplot(fig_data, aes(x = TSNE1, y = TSNE2, color = ribo_pct)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")

ggplot(fig_data, aes(x = UMAP1, y = UMAP2, color = log10_total_counts)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")

ggplot(fig_data, aes(x = UMAP1, y = UMAP2, color = log10_total_features_by_counts)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")

ggplot(fig_data, aes(x = UMAP1, y = UMAP2, color = ribo_pct)) +
  geom_point() + scale_color_continuous(low = "blue", high = "yellow")

cors <- lapply(1:10, function(i) abs(cor(W[,i], qc, method="spearman")))
cors <- unlist(cors)
bars <- data.frame(AbsoluteCorrelation=cors,
                   QC=factor(rep(colnames(qc), 10), levels=colnames(qc)),
                   Dimension=as.factor(rep(paste0("W", 1:10), each=ncol(qc))))

bars %>%
  ggplot(aes(Dimension, AbsoluteCorrelation, group=QC, fill=QC)) +
  geom_bar(stat="identity", position='dodge') +
  scale_fill_manual(values=bigPalette) + ylim(0, 1) + 
  ggtitle("Correlation between QC and W")
```

# Clustering using ZINB-WaVE projection

## Seurat

```{r seurat, results="hide", dependson="zinbwave"}
dat <- assay(filtered)
rownames(dat) <- rowData(filtered)$Symbol
rownames(dat)[rownames(dat) == "Smim20"] <- paste0("Smim20", 1:2)
colnames(dat) <- paste0("Sample", seq_len(NCOL(dat)))

seu <- CreateSeuratObject(raw.data = dat)

seu <- SetDimReduction(object = seu, reduction.type = "zinbwave", slot = "cell.embeddings", 
    new.data = W)
seu <- SetDimReduction(object = seu, reduction.type = "zinbwave", slot = "key", 
    new.data = "zinbwave")

seu <- FindClusters(object = seu, reduction.type = "zinbwave", dims.use = 1:10, 
    resolution = seq(0, 2, 0.2), print.output = 0, save.SNN = TRUE)

clustree(seu)
```

```{r clX, dependson="seurat"}
colidx <- grep("res.", colnames(seu@meta.data), fixed = TRUE)
clustMat <- as.matrix(cbind(seu@ident,
                  seu@meta.data[,colidx[-c(1, length(colidx))]]))
colnames(clustMat) <- c("Seurat", paste0("Seurat", seq(0.2, 1.8, by=.2)))

cl2 <- ClusterExperiment(norm, clustMat, transformation = log1p)

clusterLegend(cl2)[[1]][,"name"] <- paste0("S", clusterLegend(cl2)[[1]][,"name"]) 
clusterLegend(cl2)[[1]][,"color"] <- massivePalette[-(1:11)][1:NROW(clusterLegend(cl2)[[1]])]
clusterLegend(cl2)[[1]][,"color"] <- clusterLegend(cl2)[[1]][,"color"][c(7, 2, 12, 6, 8, 13, 15, 1, 5, 4, 3, 10, 11, 14, 9)]
colData(cl2) <- colData(filtered)

gfp <- rep("no", NCOL(filtered))
gfp[filtered$Batch == "TdT+GFP+"] <- "GFP"
gfp <- as.factor(gfp)
colData(cl2)$gfp <- gfp

plotClusters(cl2, whichClusters = 1, existingColors = "firstOnly",
             colData = c("Batch", "gfp"))

plotClusters(cl2, existingColors = "firstOnly")
```

## Visualization

```{r visualize_seurat, dependson="clX"}
cols1 <- clusterLegend(cl2)[[1]][, "color"]
names(cols1) <- clusterLegend(cl2)[[1]][, "name"]

pal2 <- c("#33A02C", "magenta2")

colMerged1 <- cols1[primaryClusterNamed(cl2)]

lab_s <- factor(primaryClusterNamed(cl2),
                       levels=clusterLegend(cl2)[[1]][, "name"])

plot(W, pch=19, col=colMerged1, main="W, Seurat")
legend("bottomright", levels(lab_s), fill=cols1[levels(lab_s)])

plot(tsne_zinb$Y, pch=19, col=colMerged1, main="t-SNE, All genes, Seurat",
     xlab="Dim 1", ylab="Dim 2")

plot(tsne_zinb$Y, pch=19, col=pal2[cl2$gfp])
legend("topleft", levels(cl2$gfp), fill=pal2)

plot(umap_zinb, pch=19, col=colMerged1, main="t-SNE, All genes, Seurat",
     xlab="Dim 1", ylab="Dim 2")

plot(umap_zinb, pch=19, col=pal2[cl2$gfp])
legend("bottomright", levels(cl2$gfp), fill=pal2)
```

# Marker genes

```{r heatmap_zinb, dependson=c("clX", "visualize_seurat")}
reducedDim(cl2, "zinbwave") <- W
rownames(cl2) <- rowData(filtered)$Symbol

cl2 <- makeDendrogram(cl2, reduceMethod = "zinbwave", nDims = 10)
plotDendrogram(cl2, plotType = "name", leafType = "clusters")

merged <- mergeClusters(cl2, mergeMethod = "adjP", cutoff = 0.05, plotInfo="mergeMethod", DEMethod = "limma")

plotClustersTable(merged, whichClusters = 1:2)

tab <- table(primaryClusterNamed(merged), gfp)
print(tab)
print(round(tab/rowSums(tab), 2))

merged <- makeDendrogram(merged, reduceMethod = "zinbwave", nDims = 10)

genes <- getBestFeatures(merged, contrastType = "Dendro", DEMethod="limma-voom", number = 100)
head(genes)

plotHeatmap(merged, clusterSamplesData = "dendrogramValue",
            colData = c("gfp", "Batch"),
            clusterFeaturesData=unique(genes[,"IndexInOriginal"]), breaks=.99)

top <- getBestFeatures(merged, contrastType="Dendro", DEMethod="limma", number=Inf, p.value=0.05)

xx <- factor(primaryClusterNamed(merged), 
             levels=paste0("m", 1:(nClusters(merged)[1])))
means <- t(apply(log1p(assay(merged))[top$Feature,], 1, tapply, xx, mean))

retval <- cbind(top, means[top$Feature,])[, -c(3, 5, 7, 8, 11)]
write.table(retval, "DE_genes_dendro_merged_zinb_all_batches.txt", quote=FALSE, sep="\t")

top <- getBestFeatures(merged, contrastType = "OneAgainstAll", DEMethod="limma", number=Inf, p.value=0.05)

means <- t(apply(log1p(assay(merged))[top$Feature,], 1, tapply, xx, mean))
retval <- cbind(top, means[top$Feature,])[, -c(3, 5, 7, 8, 11)]

map <- data.frame(ids = paste0("Cl", stringr::str_pad(1:nClusters(merged)[1], width=2, pad=0)),
                  Contrast = clusterLegend(merged)[[1]][,"name"], stringsAsFactors = FALSE)
retval <- dplyr::inner_join(retval, map, by = c("ContrastName" = "ids"))[,-(1:2)]

write.table(retval, "DE_genes_1vall_merged_zinb_all_batches.txt", quote=FALSE, sep="\t")
```

```{r save, dependson="clX"}
save(merged, file="data/clusterexperiment_all_batches.rda")
```

```{r allen, dependson="heatmap_zinb"}
fig2 <- c("Sst", "Chodl", "Nos1", "Etv1", "Il1rapl2", "Myh8", "Chrna2",
          "Tac2", "Crhr2", "Calb2", "Hpse", "C1ql3", "Crh", "Nts", "Pvalb",
          "Gabrg1", "Th", "Prdm8", "Calb1", "Reln", "Gpr149", "Cpne5",
          "Vipr2", "Nkx2.1")

plotHeatmap(merged, clusterSamplesData = "dendrogramValue",
            sampleData = c("gfp"),
            clusterFeaturesData=which(rownames(cl2) %in% fig2), breaks=.99)

rownames(lognorm) <- rownames(merged)

df <- data.frame(t(lognorm[intersect(c(fig2, "GFP", "tdTomato"), rownames(lognorm)),]),
                 Cluster = primaryClusterNamed(merged), Batch = cl2$Batch)

ggplot(df, aes(x = Calb1, y = Batch, fill = Batch)) +
  geom_density_ridges() + 
  scale_fill_manual(values = bigPalette) +
  ggtitle("Calb1 expression")

ggplot(df, aes(x = Calb2, y = Batch, fill = Batch)) +
  geom_density_ridges() + 
  scale_fill_manual(values = bigPalette) +
  ggtitle("Calb2 expression")

ggplot(df, aes(x = Sst, y = Batch, fill = Batch)) +
  geom_density_ridges() + 
  scale_fill_manual(values = bigPalette) +
  ggtitle("Sst expression")

ggplot(df, aes(x = Calb1, y = Cluster, fill = Cluster)) +
  geom_density_ridges() + 
  scale_fill_manual(values = cols1) +
  ggtitle("Calb1 expression")

ggplot(df, aes(x = Calb2, y = Cluster, fill = Cluster)) +
  geom_density_ridges() + 
  scale_fill_manual(values = cols1) +
  ggtitle("Calb2 expression")

ggplot(df, aes(x = Sst, y = Cluster, fill = Cluster)) +
  geom_density_ridges() + 
  scale_fill_manual(values = cols1) +
  ggtitle("Sst expression")

ggplot(df, aes(x = Pvalb, y = Cluster, fill = Cluster)) +
  geom_density_ridges() + 
  scale_fill_manual(values = cols1) +
  ggtitle("Pvalb expression")

ggplot(df, aes(x = Reln, y = Cluster, fill = Cluster)) +
  geom_density_ridges() + 
  scale_fill_manual(values = cols1) +
  ggtitle("Reln expression")

ggplot(df, aes(x = Myh8, y = Cluster, fill = Cluster)) +
  geom_density_ridges() + 
  scale_fill_manual(values = cols1) +
  ggtitle("Myh8 expression")

ggplot(df, aes(x = C1ql3, y = Cluster, fill = Cluster)) +
  geom_density_ridges() + 
  scale_fill_manual(values = cols1) +
  ggtitle("C1ql3 expression")

ggplot(df, aes(x = GFP, y = Cluster, fill = Cluster)) +
  geom_density_ridges() + 
  scale_fill_manual(values = cols1) +
  ggtitle("GFP expression")

ggplot(df, aes(x = tdTomato, y = Cluster, fill = Cluster)) +
  geom_density_ridges() + 
  scale_fill_manual(values = cols1) +
  ggtitle("tdTomato expression")
```

## Subsample analysis (cluster stability)

```{r sub_seurat}
## adapted from Allen (https://github.com/AllenInstitute/scrattch.hicat/blob/master/R/cluster.R)
jaccard <- function(m) {
  library(Matrix)
  
  ## common values:
  A <-  m %*% t(m)
  
  ## indexes for non-zero common values
  im <- Matrix::which(A > 0, arr.ind=TRUE)
  
  ## counts for each row
  b <- Matrix::rowSums(m)  
  
  ## only non-zero values of common
  Aim <- A[im]
  
  ## Jacard formula: #common / (#i + #j - #common)
  J <- sparseMatrix(
    i = im[,1],
    j = im[,2],
    x = Aim / (b[im[,1]] + b[im[,2]] - Aim),
    dims = dim(A)
    )  
  return(J)
}

pass_louvain <- function(mod.sc, adj.mat)
{
  library(Matrix)
  p <- mean(Matrix::colSums(adj.mat > 0) - 1) / nrow(adj.mat)
  n <- ncol(adj.mat)
  rand.mod1 <- 0.97 * sqrt((1 - p) / (p * n))
  rand.mod2 <- (1 - 2 / sqrt(n)) * (2 / (p * n)) ^ (2 / 3)
  rand.mod.max <- max(rand.mod1, rand.mod2, na.rm = TRUE)
  #cat("Modularity:",mod.sc, "threshold:",rand.mod.max, "\n")
  return(mod.sc > rand.mod.max)
}

jaccard_louvain.FNN <- function(x, 
                                k = 10, 
                                knn.matrix = NULL, ...)
{
  library(igraph)
  library(matrixStats)
  library(FNN)  
  
  if(is.null(knn.matrix)){
    knn.result <- FNN::get.knn(t(x), k)
    knn.matrix <- knn.result[[1]]
  }
  
  p <- as.vector(t(knn.matrix))
  
  edge <- cbind(rep(1:nrow(knn.matrix), rep(k, nrow(knn.matrix))), p)
  
  edge.unique <- cbind(rowMins(edge), rowMaxs(edge))
  edge.unique <- unique(edge.unique)
  
  knn.gr <- igraph::graph(t(edge))
  knn.matrix <- igraph::get.adjacency(knn.gr)
  
  jaccard.adj  <- jaccard(knn.matrix)
  jaccard.gr <- igraph::graph.adjacency(jaccard.adj, 
                                        mode = "undirected", 
                                        weighted = TRUE)
  
  louvain.result <- igraph::cluster_louvain(jaccard.gr)
  
  mod.sc <- igraph::modularity(louvain.result)
  
  if(pass_louvain(mod.sc, jaccard.adj)) {
    
    cl <- setNames(louvain.result$membership, row.names(x))
    
    return(cl)
    
  } else{
    
    return(NULL)
    
  }
}

internalFunctionCheck(jaccard_louvain.FNN, inputType = "X", algorithmType = "K",
                      outputType="vector")

SNN <- ClusterFunction(jaccard_louvain.FNN, inputType = "X", algorithmType = "K",
                     outputType="vector")

set.seed(282582)
cl_sub <- clusterSingle(
  as(cl2, "SingleCellExperiment"), reduceMethod = "zinbwave", nDims = 10,
  subsample = TRUE, sequential = FALSE,
  mainClusterArgs = list(clusterFunction = "hierarchical01",
                         clusterArgs = list(alpha = 0.2),
                         minSize = 10),
  subsampleArgs = list(resamp.num = 100,
                       samp.p = 0.7,
                       clusterFunction = SNN,
                       clusterArgs = list(k = 30),
                       ncores = 1)
)
cl_sub

table(primaryClusterNamed(cl_sub), primaryClusterNamed(cl2))
cl_sub <- addClusterings(cl_sub, primaryClusterNamed(cl2), clusterLabels = "Seurat") 
cl_sub <- addClusterings(cl_sub, cl2$gfp, clusterLabels = "gfp") 
clusterLabels(cl_sub)[1] <- "subsampling"
clusterLegend(cl_sub)$Seurat <- clusterLegend(cl2)$Seurat
clusterLegend(cl_sub)$gfp[,"color"] <- pal2

plotClustersTable(cl_sub, whichClusters = c("subsampling", "Seurat"))

png("co_clustering.png", width = 800, height = 800)
plotCoClustering(cl_sub, whichClusters = c("subsampling", "Seurat", "gfp"))
dev.off()
```

# Remove contaminants: S12, S13, S14

```{r remove_samples, dependson="heatmap_zinb"}
wh_rm <- which(primaryClusterNamed(cl2) %in% paste0("S", 12:14))

tsne_sub <- Rtsne(W[-wh_rm,], pca = FALSE, max_iter=5000)
umap_sub <- umap(W[-wh_rm,], n_neighbors = 15)

plot(tsne_sub$Y, pch=19, col=colMerged1[-wh_rm], xlab="Dim 1", ylab="Dim 2")
plot(tsne_sub$Y, pch=19, col=pal2[cl2$gfp[-wh_rm]], xlab="Dim 1", ylab="Dim 2")

plot(umap_sub, pch=19, col=colMerged1[-wh_rm], xlab="Dim 1", ylab="Dim 2")
plot(umap_sub, pch=19, col=pal2[cl2$gfp[-wh_rm]], xlab="Dim 1", ylab="Dim 2")
```

## Heatmap with markers

```{r heat_clust}
mgenes <- c("Calb1", "Pld5", "Tacr1", "Hpse", "Crh", "Calb2")

cl3 <- merged[,-wh_rm]
cl3 <- makeDendrogram(cl3, reduceMethod = "zinbwave", nDims = 10,
                      whichCluster = "Seurat")
plotDendrogram(cl3, plotType = "name", leafType = "clusters")

cl3 <- addClusterings(cl3, cl3$gfp, clusterLabels = "gfp")
clusterLegend(cl3)$gfp[,"color"] <- pal2

cl3 <- plotClusters(cl3, whichClusters = c("Seurat", "mergeClusters"),
             existingColors = "firstOnly", resetColors = TRUE)

plotHeatmap(cl3, clusterSamplesData = "dendrogramValue",
            clusterFeaturesData=which(rownames(cl3) %in% mgenes),
            breaks=.9,
            whichClusters = c("Seurat", "mergeClusters", "gfp"))

plotHeatmap(cl3, clusterSamplesData = "dendrogramValue",
            clusterFeaturesData=which(rownames(cl3) %in% mgenes),
            breaks=.9,
            whichClusters = c("mergeClusters", "gfp"))

pdf("Figure7a.pdf")
plotHeatmap(cl3, clusterSamplesData = "dendrogramValue",
            clusterFeaturesData=which(rownames(cl3) %in% mgenes),
            breaks=.9,
            whichClusters = c("mergeClusters", "gfp"))
dev.off()

df <- data.frame(t(assay(cl3)[mgenes,]), gfp = colData(cl3)$gfp,
                 Cluster = factor(primaryClusterNamed(cl3), levels=paste0("m", c(1:4, 8:13))))

cols1 <- clusterLegend(cl3)[[1]][, "color"]
names(cols1) <- clusterLegend(cl3)[[1]][, "name"]
colMerged3 <- cols1[primaryClusterNamed(cl3)]

plot(assay(cl3)["Calb2",], assay(cl3)["Hpse",], pch=19, col=colMerged3, xlab="Calb2", ylab="Hpse")
plot(assay(cl3)["Calb2",], assay(cl3)["Hpse",], pch=19, col=pal2[cl3$gfp], xlab="Calb2", ylab="Hpse")
```

# Allen clusters

```{r allen_clusters, cache=FALSE}
load("data/allen.rda")
sce2 <- as(allen[, grep("^Sst", allen$cluster)], "SingleCellExperiment")
rowData(sce2)$feature_symbol <- rownames(sce2)
assayNames(sce2) <- "counts"
genes <- intersect(rownames(sce2), rownames(cl2))
sce2 <- scater::normalize(sce2[genes,])
sce2 <- scater::runPCA(sce2, ncomponents = 30, method = "irlba",
                       ntop = 1000)
rm(allen)

sce <- ClusterExperiment(sce2, colData(sce2)$cluster)

sce <- makeDendrogram(sce, reduceMethod = "PCA", nDims = 30)
plotDendrogram(sce, plotType = "name", leafType = "clusters")

plotHeatmap(sce, clusterSamplesData = "dendrogramValue",
            clusterFeaturesData=which(rownames(sce) %in% mgenes),
            breaks=.8)

pal_allen <- clusterLegend(sce)[[1]][,"color"]
names(pal_allen) <- clusterLegend(sce)[[1]][,"name"]

plot(counts(sce)["Calb2",], counts(sce)["Hpse",], pch=19, col=pal_allen[primaryClusterNamed(sce)], xlab="Calb2", ylab="Hpse")
plotClusterLegend(sce)

boxplot(logcounts(sce)["Hpse",]~primaryClusterNamed(sce), col=pal_allen[primaryClusterNamed(sce)], las=3)
```

## Mapping clusters with `scmap`

```{r scmap, cache=FALSE, results="asis"}
sce2 <- selectFeatures(sce2, suppress_plot = TRUE)
sce2 <- indexCluster(sce2, cluster_col = "cluster")

cl4 <- as(cl3, "SingleCellExperiment")
rowData(cl4)$feature_symbol <- rownames(cl4)
normcounts(cl4) <- assay(cl4)
logcounts(cl4) <- log2(normcounts(cl4) + 1)

scmapCluster_results <- scmapCluster(
  projection = cl4, 
  index_list = list(
    Allen = metadata(sce2)$scmap_cluster_index
  ), threshold = 0.4 #0.6
)

cols1 <- clusterLegend(cl3)[[1]][,"color"]
names(cols1) <- clusterLegend(cl3)[[1]][,"name"]
cols <- sapply(cols1, function(x) {
  y <- col2rgb(x)[,1]
  rgb(y[1]/255, y[2]/255, y[3]/255)
})
# 
# plot(
#   getSankey(
#     primaryClusterNamed(cl3), 
#     scmapCluster_results$scmap_cluster_labs[,'Allen'],
#     plot_height = 1000, colors = cols
#   ), tag="chart"
# )
# 
# plot(
#   getSankey(
#     cl4$gfp, 
#     scmapCluster_results$scmap_cluster_labs[,'Allen'],
#     plot_height = 1000, colors = c("grey", "green")
#   ), tag="chart"
# )

```

```{r scmap_table, dependson = "scmap"}
mat <- as.matrix(table(scmapCluster_results$scmap_cluster_labs, primaryClusterNamed(cl3)))
print(data.frame(Our_cluster = colnames(mat),
                 AIBS = rownames(mat)[apply(mat, 2, which.max)]))
```

```{r static, eval = FALSE}
library(ggforce)
library(dplyr)
alpha <- 0.7

dat_raw <- data.frame("Ref_Allen" = scmapCluster_results$scmap_cluster_labs[,1],
                      "Clusters" = primaryClusterNamed(cl3),
                      stringsAsFactors = FALSE)

dat <- dat_raw %>%
  dplyr::filter(! Clusters %in% c("S12", "S13", "S14")) %>%
  group_by(Clusters, Ref_Allen) %>%
  summarise(freq = n()) %>%
  ungroup()

dat_ggforce <- dat  %>%
  gather_set_data(2:1) %>%        # <- ggforce helper function
  dplyr::arrange(x, Ref_Allen, desc(Clusters))

dat_filtered <- dat_ggforce %>%
  dplyr::filter(freq >= 10)

ggplot(dat_filtered, aes(x = x, id = id, split = y, value = freq)) +
  geom_parallel_sets(aes(fill = Clusters), alpha = alpha, axis.width = 0.2,
                     n=100, strength = 0.5) +
  geom_parallel_sets_axes(axis.width = 0.25, fill = "gray95",
                          color = "gray80", size = 0.15) +
  geom_parallel_sets_labels(colour = 'gray35', size = 4.5, angle = 0, fontface="bold") +
  scale_fill_manual(values  = cols) +
  scale_color_manual(values = cols) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 20, face = "bold"),
    axis.title.x  = element_blank()
    ) -> p1
p1

ggsave("riverplot.pdf", p1, width = 10, height = 10)

dat_raw <- data.frame("Ref_Allen" = scmapCluster_results$scmap_cluster_labs[,1],
                      "Clusters" = primaryClusterNamed(cl3),
                      "GFP" = cl3$gfp,
                      stringsAsFactors = FALSE)

dat <- dat_raw %>%
  dplyr::filter(! Clusters %in% c("S12", "S13", "S14")) %>%
  group_by(GFP, Ref_Allen) %>%
  summarise(freq = n()) %>%
  ungroup()

dat_ggforce <- dat  %>%
  gather_set_data(2:1) %>%        # <- ggforce helper function
  dplyr::arrange(x, Ref_Allen, desc(GFP))

dat_filtered <- dat_ggforce %>%
  dplyr::filter(freq >= 10)

cols2 <- c("green", "grey")

ggplot(dat_filtered, aes(x = x, id = id, split = y, value = freq)) +
  geom_parallel_sets(aes(fill = GFP), alpha = alpha, axis.width = 0.2,
                     n=100, strength = 0.5) +
  geom_parallel_sets_axes(axis.width = 0.25, fill = "gray95",
                          color = "gray80", size = 0.15) +
  geom_parallel_sets_labels(colour = 'gray35', size = 4.5, angle = 0, fontface="bold") +
  scale_fill_manual(values  = cols2) +
  scale_color_manual(values = cols2) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 20, face = "bold"),
    axis.title.x  = element_blank()
    ) -> p2
p2
ggsave("riverplot_gfp.pdf", p2, width = 10, height = 10)
```

# Session Info

```{r}
sessionInfo()
```
